Modelling Prehistoric Iconographic Compositions. The \proglang{R} package \pkg{iconr}

### 2020-12-18 article_rvTH17.Rnw
article

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{Modelling Prehistoric Iconographic Compositions. The \proglang{R} package \pkg{decorr}}
\Plaintitle{Modelling Prehistoric Iconographic Compositions. The R package 'decorr'}
\Shorttitle{Modelling Prehistoric Iconographic Compositions}

%% - \Abstract{} almost as usual
\Abstract{
By definition, Prehistorical societies are characterised by the absence of a writing system. During, the largest part of human history -- from far -- symbolic expressions belong to illiterate societies which express themself with rock-art paintings, pottery decorations, figurines and statuary, etc., and a lot of now disappeared carved woods, textil design, etc. At the composition level, recognition of meaningful associations of signs and reccurent patterns indicate clearly the existence of social conventions in the way to display and to read these expressions.
We present the \pkg{decorr} \proglang{R} package which grounds concepts, methods and tools to analyse any ancient graphical systems. Our asumption is that i) any graphical system is a spatial distribution of features, and ii) these features have possibly any meaningful relationships one with another depending on their pairwise spatial proximities. To model the graphical compositions we employ concepts coming from the Graph Theory. To ensure the feasability of this type of analysis, we propose a GIS-based method for inputs and a serie of functions for data management.
}

@@

In this article we present the \proglang{R} package \pkg{decorr}. Its purpose is to formalise a method and offer tools to analyse any graphical content (here: decorations) with methods coming from raph Theory, GIS and binary topological relationships (\emph{birel}) modeling. This package has been grounded on the seminal work of \citet{Alexander08} and its first IT implementation by \citet{Huet18a}.

@@

<<dec.1.4, echo=FALSE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=15, fig.asp=0.5630631, fig.cap="\\label{fig:dec.1.4} Cerro Muriano 1 and Zarza de Montanchez decorations.">>=
library(decorr)
par(mar=c(0, 0, 0, 0))
nds.df <- read_nds(site = "Cerro Muriano", 
                   decor = "Cerro Muriano 1",
                   doss = system.file("extdata", 
                                      package = "decorr"))
eds.df <- read_eds(site = "Cerro Muriano",
                   decor = "Cerro Muriano 1",
                   doss = system.file("extdata", 
                                      package = "decorr"))
cm1 <- plot_dec_grph(nodes = nds.df,
                     edges = eds.df,
                     site = "Cerro Muriano",
                     decor = "Cerro Muriano 1",
                     nd.var = "type",
                     lbl.size = 1.7,
                     doss = system.file("extdata", 
                                        package = "decorr"))
cm1.img <- magick::image_read(cm1)
nds.df <- read_nds(site = "Zarza de Montanchez", 
                   decor = "Zarza De Montanchez",
                   doss = system.file("extdata", 
                                      package = "decorr"))
eds.df <- read_eds(site = "Zarza de Montanchez",
                   decor = "Zarza De Montanchez",
                   doss = system.file("extdata", 
                                      package = "decorr"))
zm <- plot_dec_grph(nodes = nds.df,
                            edges = eds.df,
                            site = "Zarza de Montanchez",
                            decor = "Zarza De Montanchez",
                            nd.var = "type",
                            lbl.size = 1.7,
                            doss = system.file("extdata", 
                                               package = "decorr"))
zm.img <- magick::image_read(zm)
out.img <- magick::image_append(c(cm1.img, zm.img))
plot(out.img)
@

@@

Nodes are uniquely identified by a number. But this does not identify or classify the nodes by the corresponding GU meaning or type. For this, name nodes could be assigned. However this presents two problems: nodes with repeated GU type in the same graph could not be handled and comparisons by different GU classifications would be hindered. For this reason, the R package decorr considers that the GU type is assigned in a node attribute or variable. Thus, at least one node variable must be defined in order to analyse and compare iconographical compositions. By default this node variable is named \code{type}. In this article examples, we consider node \code{type} values ('anthropomorph', 'sword', etc.), but the package user can add other variables (\code{technique}, \code{color}, etc.), which would produce complementary analyses.

@@

\section{The R package decorr} \label{sec:models}

<<results='asis', include=FALSE, echo=FALSE, warning=FALSE>>=
library("decorr")
library("igraph")
library("magick")
@
%
The \pkg{decorr} package can be downloaded from GitHub
%
\begin{CodeChunk}
\begin{CodeInput}
R>   devtools::install_github("zoometh/iconr", build_vignettes=TRUE)
\end{CodeInput}
\end{CodeChunk}
%

@@

The \pkg{decorr} package imports the following packages:
\begin{itemize}
\setlength\itemsep{.1em}
\item \pkg{magick} for image manipulation \citep{Ooms18}
\item \pkg{igraph} for graph and network analysis \citep{Csardi06}
\item \pkg{rgdal}  to read \emph{shapefiles} of nodes and edges \citep{Bivand19}
\item \pkg{grDevices} for colors and font plotting, \pkg{graphics} for graphics, \pkg{utils} and \pkg{methods} for formally defined methods and \emph{varia} methods (all combinations, etc.) \citep{R19}
\end{itemize}

@@

A small testing dataset is provided with the \pkg{decorr} package, stored in its \code{extdata} folder. It contains dataframes of nodes' coordinates, edges' coordinates, decorations' listing in different formats (\code{.tsv}, \code{.csv} or \emph{shapefiles}), and the decoration drawings (\code{.jpg}) of five Late Bronze Age stelae belonging to the so-called 'Warrior stelae' family -- with about 140 stelae -- mostly concentrated in the Southwest Iberian peninsula \citep{Celestino01}.

\subsubsection{Imgs dataframe}
The inventory of decorations is stored in the comma-separated values file \code{imgs.csv} and the tab-separated values files \code{imgs.tsv}. Any of them can be read as a dataframe (Table \ref{tab:timg}),
%
<<results='asis', echo=TRUE, warning=FALSE>>=
imgs <- read.table(system.file("extdata", "imgs.tsv", package = "decorr"),
                   sep="\t", stringsAsFactors = FALSE)
@
The field \code{imgs$idf} is the short name of the decoration, useful during statistical analysis. The primary key of each decoration is the concatenate of \code{imgs$site} and \code{imgs$decor}. These keys will allow joints with the other dataframes (nodes and edges).
<<results='asis', echo=FALSE, warning=FALSE>>=
print(xtable::xtable(imgs,
                     caption = "The studied corpus, the \\code{imgs.tsv} dataframe",
                     label = "tab:timg",
                     size=6,
                     align=c(
                       "p{0.01cm}",
                       "c",
                       ">{\\raggedright\\arraybackslash\\rule{0pt}{1em}\\addtolength{\\baselineskip}{-1.5pt}}p{2.1cm}",
                       ">{\\raggedright\\arraybackslash\\addtolength{\\baselineskip}{-1.5pt}}p{2.1cm}",
                       "p{8.5cm}")),
      table.placement="!htbp",
      include.rownames=FALSE)
@

At first the drawings dataset \citep{DiazGuardamino10} can be checked by using the information in the \code{imgs} dataframe and the \pkg{magick} package (Figure \ref{fig:dataset})
%
<<timg, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=15, fig.asp=0.7507508, fig.cap="\\label{fig:dataset}Decoration images of the training dataset">>=
pth <- system.file("extdata", package = "decorr")
lims <- list()
for(i in 1:nrow(imgs)){
  i1 <- image_read(paste0(pth,"/",imgs$img[i]))
  lbl.txt <- paste0(paste(imgs[i,], collapse = "\n"), "\n",
                    image_info(i1)$width, "*", image_info(i1)$height, " px")
  lims[[i]] <- image_annotate(i1, lbl.txt, location = "northwest",
                              size = 28, color = "red")
}
out.img <- image_append(c(image_append(c(lims[[1]], lims[[2]], lims[[3]])),
                          image_append(c(lims[[4]], lims[[5]]))),
                        stack = TRUE)
par(mar=c(0, 0, 0, 0))
plot(out.img)
@
%
To construct a graph overlapping the decoration images listed in the \code{imgs} dataframe, the first step is to load nodes and edges dataframes.
% \begin{itemize}
% \item The \code{nodes} dataframe
% \end{itemize}

@@

<<results='asis', echo=TRUE>>=
nodes <- read.table(system.file("extdata", "nodes.csv", package = "decorr"),
                    sep=";", stringsAsFactors = FALSE)
@

@@

\subsubsection{Edges dataframe}
Analogously, the edges for the five decorations are stored in different formats in \code{edges.csv}, \code{edges.tsv}, and \code{edges.shp}. This can be read by 
<<results='asis', echo=TRUE>>=
edges <- read.table(system.file("extdata", "edges.csv", 
                                package = "decorr"),
                    sep=";",stringsAsFactors = FALSE)
@

@@

\bigskip
The \pkg{decorr} package functions can be divided into:
\begin{enumerate}
\item management functions
\item graphical functions
\item single decoration functions
\item comparisons functions
\end{enumerate}

@@

<<igraphs, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=6, fig.height=4, fig.cap="\\label{fig:igraph}R interface. Plot of the first graph of the list (Cerro Muriano 1 decoration) with graph drawing on x, y coordinates (left), and force-directed graph drawing with nodes sized on their degree centralities (right)">>=
lgrph <- list_dec(imgs, nodes, edges)
g <- lgrph[[1]]
par(mar = c(0, 0, 0, 0), mfrow = c(1, 2))
coords <- layout.fruchterman.reingold(g)
plot(g,
     vertex.size = 14,
     vertex.label.family = "Helvetica",
     edge.color = "orange",
     vertex.label.cex = .8
)
plot(g,
     layout = coords,
     vertex.size = 4 + (degree(g)*10),
     vertex.label.family = "Helvetica",
     edge.color = "orange",
     vertex.label.cex = .8
)
@

@@

<<ibah, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=3.5, fig.asp=1.126126, fig.cap="\\label{fig:ibah}Ibahernando decoration, node 1 overlaps node 2 and node 3">>=
par(mar=c(0,0,0,0))
nds.df <- read_nds(site = "Ibahernando", 
                   decor = "Ibahernando",
                   doss = system.file("extdata", 
                                      package = "decorr"))
eds.df <- read_eds(site = "Ibahernando",
                   decor = "Ibahernando",
                   doss = system.file("extdata", 
                                      package = "decorr"))
img.graph <- plot_dec_grph(nodes = nds.df,
                           edges = eds.df,
                           site = "Ibahernando",
                           decor = "Ibahernando",
                           lbl.size = 2,
                           doss = system.file("extdata", 
                                              package = "decorr"))
plot(magick::image_read(img.graph))
@

@@

<<ibah2, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=8, fig.asp=0.5630631, fig.cap="\\label{fig:ibah2}Ibahernando stelae before and after the selection of node 4 (sword) graph component">>=
par(mar=c(0,0,0,0))
selected.nd <- 4
nds.df <- read_nds(site = "Ibahernando", 
                   decor = "Ibahernando",
                   doss = system.file("extdata", 
                                      package = "decorr"))
eds.df <- read_eds(site = "Ibahernando",
                   decor = "Ibahernando",
                   doss = system.file("extdata", 
                                      package = "decorr"))
l_dec_df <- contemp_nds(nds.df, eds.df, selected.nd)
Ibahernando <- plot_dec_grph(nodes = nds.df,
                             edges = eds.df,
                             site = "Ibahernando",
                             decor = "Ibahernando",
                             nd.var = "type",
                             lbl.color = "brown",
                             lbl.size=2.2,
                             doss = system.file("extdata", 
                                                package = "decorr"))
Ibahernando.img <- magick::image_read(Ibahernando)
Ibahernando.contemp <- plot_dec_grph(nodes = l_dec_df[[1]],
                                     edges = l_dec_df[[2]],
                                     site = "Ibahernando",
                                     decor = "Ibahernando",
                                     nd.var = "type",
                                     lbl.color = "brown",
                                     lbl.size=2.2,
                                     doss = system.file("extdata", 
                                                        package = "decorr"))
Ibahernando.contemp.img <- magick::image_read(Ibahernando.contemp)
out <- magick::image_append(c(Ibahernando.img, Ibahernando.contemp.img))
plot(out)
@

@@

The \pkg{decorr} package has two graphical functions: \code{labels_shadow()} and \code{side_plot()}.

@@

<<nd, results='asis', echo=TRUE>>=
caption <- "Torrejon Rubio 1 decoration node 6: the sword."
sit <- "Torrejon Rubio" ; dec <- "Torrejon Rubio 1"
nds.df <- read_nds(site = sit, decor = dec, dev = ".shp",
                   doss = system.file("extdata", package = "decorr"))
print(xtable::xtable(nds.df[nds.df$id == 6, ],
                     caption = caption,
                     label = "tab:nd",
                     size = 7),
      include.rownames = FALSE,
      table.placement = "!htbp")
@

@@

<<ed, results='asis', echo=TRUE >>=
caption <- paste0("First edge of the Torrejon Rubio 1 decoration, ",
                  "the proximity between the sword (node 6) ",
                  "and the fibula (node 5)")
eds.df <- read_eds(site = sit, decor = dec, dev = ".tsv",
                   doss = system.file("extdata", package = "decorr"))
print(xtable::xtable(eds.df[1,],
                     caption = caption,
                     label = "tab:ed",
                     size = 7),
      include.rownames = FALSE,
      table.placement = "!htbp")
@

@@

<<trst, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=3.5, fig.asp=1.126126, fig.cap="\\label{fig:trst}Torrejon Rubio 1 decoration">>=
par(mar=c(0, 0, 0, 0))
sit <- "Torrejon Rubio" ; dec <- "Torrejon Rubio 1"
nds.df <- read_nds(site = sit, decor = dec, dev = ".tsv",
                   doss = system.file("extdata", package = "decorr"))
eds.df <- read_eds(site = sit, decor = dec, dev = ".tsv",
                   doss = system.file("extdata", package = "decorr"))
img.graph <- plot_dec_grph(nodes = nds.df,
                           edges = eds.df,
                           site = sit,
                           decor = dec,
                           doss = system.file("extdata", 
                                              package = "decorr"),
                           nd.var = "type", lbl.size = 2)
out <- magick::image_read(img.graph)
plot(out)
@

@@

<<echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.asp=0.5630631, fig.cap="\\label{fig:ceds}Comparison between decoration 1 (Cerro Muriano 1) and decoration 4 (Zarza de Montsanchez)">>=
par(mar=c(0,0,0,0))
g.compar <- list_compar(lgrph, "type")
eds_compar <- plot_compar(g.compar, c(1, 4),
                          focus = "edges",
                          doss = system.file("extdata", package = "decorr"))
plot(magick::image_trim(magick::image_read(eds_compar)))
@

@@

<<results='asis', echo=TRUE, fig.height= 5>>=
dataDir <- system.file("extdata", package = "decorr")
nds.df <- read_nds(site = "Ibahernando",
                   decor = "Ibahernando",
                   doss = dataDir)
eds.df <- read_eds(site = "Ibahernando",
                   decor = "Ibahernando",
                   doss = dataDir)
l_dec_df <- contemp_nds(nds.df, eds.df, selected.nd = 2)
l_dec_df[[1]]$site <- l_dec_df[[1]]$decor <- NULL
l_dec_df[[2]]$site <- l_dec_df[[2]]$decor <- NULL
grph <- graph_from_data_frame(l_dec_df[[2]], vertices = l_dec_df[[1]],
                           directed = FALSE)
grph$site <- grph$decor <- "Ibahernando"
grph$name <- 5
lgrph[[5]] <- grph
@

@@

In Archaeology, the iconography has been considered as a privileged elements to the identification of cultural facies \citep{Renfrew91}. By integrating the recording, the management and the analysis process in a simple manner, within the framework of graph Theory and binary topological telationships (\emph{birel}), the \pkg{decorr} framework opens possibilities of studying graphical systems with normalized indexes over large series (Later Paleolithic cave paintings, \emph{cups-and-rings} art, Valcamonica Iron Age \emph{pittoti}, etc.) a long period of time, and on heterogeneous source data.
\\
The \pkg{decorr} framework is grounded on a parsimonious model, where Tobler's '\emph{near things}' \citep{Tobler70} of any iconographical content is recorded with a minimal of \emph{a priori} definitions. Those definitions only concern the graph elements: graphical units variables (ie, nodes variables) and types of relations they share (edges). The current edge notation with \emph{normal} edge, \emph{attribute} edge, and \emph{diachronic} edge allow to describe most cases of Prehistorical iconography, but can also be extended to accept new types like, for example, the \emph{interact} edge when two graphical units interact (ie, when an action is visible). For the nodes, we have only considered the iconographical \code{type} (\code{personnage}, \code{casque}, etc.) as the studied variable (\code{nd.var}), but the package user can create and select any other structured vocabularies: \code{technique}, \code{color}, etc. Tree structures for hierarchical datasets allow generalization processes (up, the parent level) and specification processes (down, the children level). Such a formalism can be used to weight the differences between nodes, to conduct analysis with different level of precision or to overcome idiosynchratic typologies issues. To record the relative chronology of archaeological layers (in a case of an excavation) or the relative chronology of iconographical layers (in case of a decorated support), archaeologists use quite commonly a hierachical dataset to model : the Harris matrix. The \emph{diachronic} edge, by recording the anterior-posterior relation between pairwises of graphical units, is first operator permiting to trace the Harris matrix directly of the studied decoration. The Harris matrix can be then traced with R data tree package (\pkg{data.tree}, \cite{Glur19}) and integrated into an interactive Reingold-Tilford tree diagram  (\pkg{collapsibleTree}, \cite{Khan18}).

>>>>

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{Modelling Prehistoric Iconographic Compositions. The \proglang{R} package \pkg{iconr}}
\Plaintitle{Modelling Prehistoric Iconographic Compositions. The R package 'iconr'}
\Shorttitle{Modelling Prehistoric Iconographic Compositions}

%% - \Abstract{} almost as usual
\Abstract{
By definition, Prehistorical societies are characterised by the absence of a writing system. During, the largest part of human history -- from far -- symbolic expressions belong to illiterate societies which express themself with rock-art paintings, pottery decorations, figurines and statuary, etc., and a lot of now disappeared carved woods, textil design, etc. At the composition level, recognition of meaningful associations of signs and reccurent patterns indicate clearly the existence of social conventions in the way to display and to read these expressions.
We present the \pkg{iconr} \proglang{R} package which grounds concepts, methods and tools to analyse any ancient graphical systems. Our asumption is that i) any graphical system is a spatial distribution of features, and ii) these features have possibly any meaningful relationships one with another depending on their pairwise spatial proximities. To model the graphical compositions we employ concepts coming from the Graph Theory. To ensure the feasability of this type of analysis, we propose a GIS-based method for inputs and a serie of functions for data management.
}

@@

In this article we present the \proglang{R} package \pkg{iconr}. Its purpose is to formalise a method and offer tools to analyse any graphical content (here: decorations) with methods coming from raph Theory, GIS and binary topological relationships (\emph{birel}) modeling. This package has been grounded on the seminal work of \citet{Alexander08} and its first IT implementation by \citet{Huet18a}.

@@

<<dec.1.4, echo=FALSE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=15, fig.asp=0.5630631, fig.cap="\\label{fig:dec.1.4} Cerro Muriano 1 and Zarza de Montanchez decorations.">>=
library(iconr)
par(mar=c(0, 0, 0, 0))
nds.df <- read_nds(site = "Cerro Muriano", 
                   decor = "Cerro Muriano 1",
                   doss = system.file("extdata", 
                                      package = "iconr"))
eds.df <- read_eds(site = "Cerro Muriano",
                   decor = "Cerro Muriano 1",
                   doss = system.file("extdata", 
                                      package = "iconr"))
cm1 <- plot_dec_grph(nodes = nds.df,
                     edges = eds.df,
                     site = "Cerro Muriano",
                     decor = "Cerro Muriano 1",
                     nd.var = "type",
                     lbl.size = 1.7,
                     doss = system.file("extdata", 
                                        package = "iconr"))
cm1.img <- magick::image_read(cm1)
nds.df <- read_nds(site = "Zarza de Montanchez", 
                   decor = "Zarza De Montanchez",
                   doss = system.file("extdata", 
                                      package = "iconr"))
eds.df <- read_eds(site = "Zarza de Montanchez",
                   decor = "Zarza De Montanchez",
                   doss = system.file("extdata", 
                                      package = "iconr"))
zm <- plot_dec_grph(nodes = nds.df,
                            edges = eds.df,
                            site = "Zarza de Montanchez",
                            decor = "Zarza De Montanchez",
                            nd.var = "type",
                            lbl.size = 1.7,
                            doss = system.file("extdata", 
                                               package = "iconr"))
zm.img <- magick::image_read(zm)
out.img <- magick::image_append(c(cm1.img, zm.img))
plot(out.img)
@

@@

Nodes are uniquely identified by a number. But this does not identify or classify the nodes by the corresponding GU meaning or type. For this, name nodes could be assigned. However this presents two problems: nodes with repeated GU type in the same graph could not be handled and comparisons by different GU classifications would be hindered. For this reason, the R package iconr considers that the GU type is assigned in a node attribute or variable. Thus, at least one node variable must be defined in order to analyse and compare iconographical compositions. By default this node variable is named \code{type}. In this article examples, we consider node \code{type} values ('anthropomorph', 'sword', etc.), but the package user can add other variables (\code{technique}, \code{color}, etc.), which would produce complementary analyses.

@@

\section{The R package iconr} \label{sec:models}

<<results='asis', include=FALSE, echo=FALSE, warning=FALSE>>=
library("iconr")
library("igraph")
library("magick")
@
%
The \pkg{iconr} package can be downloaded from GitHub
%
\begin{CodeChunk}
\begin{CodeInput}
R>   devtools::install_github("zoometh/iconr", build_vignettes=TRUE)
\end{CodeInput}
\end{CodeChunk}
%

@@

The \pkg{iconr} package imports the following packages:
\begin{itemize}
\setlength\itemsep{.1em}
\item \pkg{magick} for image manipulation \citep{Ooms18}
\item \pkg{igraph} for graph and network analysis \citep{Csardi06}
\item \pkg{rgdal}  to read \emph{shapefiles} of nodes and edges \citep{Bivand19}
\item \pkg{grDevices} for colors and font plotting, \pkg{graphics} for graphics, \pkg{utils} and \pkg{methods} for formally defined methods and \emph{varia} methods (all combinations, etc.) \citep{R19}
\end{itemize}

@@

A small testing dataset is provided with the \pkg{iconr} package, stored in its \code{extdata} folder. It contains dataframes of nodes' coordinates, edges' coordinates, decorations' listing in different formats (\code{.tsv}, \code{.csv} or \emph{shapefiles}), and the decoration drawings (\code{.jpg}) of five Late Bronze Age stelae belonging to the so-called 'Warrior stelae' family -- with about 140 stelae -- mostly concentrated in the Southwest Iberian peninsula \citep{Celestino01}.

\subsubsection{Imgs dataframe}
The inventory of decorations is stored in the comma-separated values file \code{imgs.csv} and the tab-separated values files \code{imgs.tsv}. Any of them can be read as a dataframe (Table \ref{tab:timg}),
%
<<results='asis', echo=TRUE, warning=FALSE>>=
imgs <- read.table(system.file("extdata", "imgs.tsv", package = "iconr"),
                   sep="\t", stringsAsFactors = FALSE)
@
The field \code{imgs$idf} is the short name of the decoration, useful during statistical analysis. The primary key of each decoration is the concatenate of \code{imgs$site} and \code{imgs$decor}. These keys will allow joints with the other dataframes (nodes and edges).
<<results='asis', echo=FALSE, warning=FALSE>>=
print(xtable::xtable(imgs,
                     caption = "The studied corpus, the \\code{imgs.tsv} dataframe",
                     label = "tab:timg",
                     size=6,
                     align=c(
                       "p{0.01cm}",
                       "c",
                       ">{\\raggedright\\arraybackslash\\rule{0pt}{1em}\\addtolength{\\baselineskip}{-1.5pt}}p{2.1cm}",
                       ">{\\raggedright\\arraybackslash\\addtolength{\\baselineskip}{-1.5pt}}p{2.1cm}",
                       "p{8.5cm}")),
      table.placement="!htbp",
      include.rownames=FALSE)
@

At first the drawings dataset \citep{DiazGuardamino10} can be checked by using the information in the \code{imgs} dataframe and the \pkg{magick} package (Figure \ref{fig:dataset})
%
<<timg, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=15, fig.asp=0.7507508, fig.cap="\\label{fig:dataset}Decoration images of the training dataset">>=
pth <- system.file("extdata", package = "iconr")
lims <- list()
for(i in 1:nrow(imgs)){
  i1 <- image_read(paste0(pth,"/",imgs$img[i]))
  lbl.txt <- paste0(paste(imgs[i,], collapse = "\n"), "\n",
                    image_info(i1)$width, "*", image_info(i1)$height, " px")
  lims[[i]] <- image_annotate(i1, lbl.txt, location = "northwest",
                              size = 28, color = "red")
}
out.img <- image_append(c(image_append(c(lims[[1]], lims[[2]], lims[[3]])),
                          image_append(c(lims[[4]], lims[[5]]))),
                        stack = TRUE)
par(mar=c(0, 0, 0, 0))
plot(out.img)
@
%
To construct a graph overlapping the decoration images listed in the \code{imgs} dataframe, the first step is to load nodes and edges dataframes.
% \begin{itemize}
% \item The \code{nodes} dataframe
% \end{itemize}

@@

<<results='asis', echo=TRUE>>=
nodes <- read.table(system.file("extdata", "nodes.csv", package = "iconr"),
                    sep=";", stringsAsFactors = FALSE)
@

@@

\subsubsection{Edges dataframe}
Analogously, the edges for the five decorations are stored in different formats in \code{edges.csv}, \code{edges.tsv}, and \code{edges.shp}. This can be read by 
<<results='asis', echo=TRUE>>=
edges <- read.table(system.file("extdata", "edges.csv", 
                                package = "iconr"),
                    sep=";",stringsAsFactors = FALSE)
@

@@

\bigskip
The \pkg{iconr} package functions can be divided into:
\begin{enumerate}
\item management functions
\item graphical functions
\item single decoration functions
\item comparisons functions
\end{enumerate}

@@

<<igraphs, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=4, fig.height=4, fig.cap="\\label{fig:igraph}R interface. Plot of the first graph of the list (Cerro Muriano 1 decoration) with graph drawing on x, y coordinates (left), and force-directed graph drawing with nodes sized on their degree centralities (right)">>=
lgrph <- list_dec(imgs, nodes, edges)
g <- lgrph[[1]]
par(mar = c(0, 0, 0, 0), mfrow = c(1, 2))
coords <- layout.fruchterman.reingold(g)
plot(g,
     vertex.size = 14,
     vertex.label.family = "Helvetica",
     edge.color = "orange",
     vertex.label.cex = .8
)
plot(g,
     layout = coords,
     vertex.size = 4 + (degree(g)*10),
     vertex.label.family = "Helvetica",
     edge.color = "orange",
     vertex.label.cex = .8
)
@

@@

<<ibah, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=3.5, fig.asp=1.126126, fig.cap="\\label{fig:ibah}Ibahernando decoration, node 1 overlaps node 2 and node 3">>=
par(mar=c(0,0,0,0))
nds.df <- read_nds(site = "Ibahernando", 
                   decor = "Ibahernando",
                   doss = system.file("extdata", 
                                      package = "iconr"))
eds.df <- read_eds(site = "Ibahernando",
                   decor = "Ibahernando",
                   doss = system.file("extdata", 
                                      package = "iconr"))
img.graph <- plot_dec_grph(nodes = nds.df,
                           edges = eds.df,
                           site = "Ibahernando",
                           decor = "Ibahernando",
                           lbl.size = 2,
                           doss = system.file("extdata", 
                                              package = "iconr"))
plot(magick::image_read(img.graph))
@

@@

<<ibah2, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=8, fig.asp=0.5630631, fig.cap="\\label{fig:ibah2}Ibahernando stelae before and after the selection of node 4 (sword) graph component">>=
par(mar=c(0,0,0,0))
selected.nd <- 4
nds.df <- read_nds(site = "Ibahernando", 
                   decor = "Ibahernando",
                   doss = system.file("extdata", 
                                      package = "iconr"))
eds.df <- read_eds(site = "Ibahernando",
                   decor = "Ibahernando",
                   doss = system.file("extdata", 
                                      package = "iconr"))
l_dec_df <- contemp_nds(nds.df, eds.df, selected.nd)
Ibahernando <- plot_dec_grph(nodes = nds.df,
                             edges = eds.df,
                             site = "Ibahernando",
                             decor = "Ibahernando",
                             nd.var = "type",
                             lbl.color = "brown",
                             lbl.size=2.2,
                             doss = system.file("extdata", 
                                                package = "iconr"))
Ibahernando.img <- magick::image_read(Ibahernando)
Ibahernando.contemp <- plot_dec_grph(nodes = l_dec_df[[1]],
                                     edges = l_dec_df[[2]],
                                     site = "Ibahernando",
                                     decor = "Ibahernando",
                                     nd.var = "type",
                                     lbl.color = "brown",
                                     lbl.size=2.2,
                                     doss = system.file("extdata", 
                                                        package = "iconr"))
Ibahernando.contemp.img <- magick::image_read(Ibahernando.contemp)
out <- magick::image_append(c(Ibahernando.img, Ibahernando.contemp.img))
plot(out)
@

@@

The \pkg{iconr} package has two graphical functions: \code{labels_shadow()} and \code{side_plot()}.

@@

<<nd, results='asis', echo=TRUE>>=
caption <- "Torrejon Rubio 1 decoration node 6: the sword."
sit <- "Torrejon Rubio" ; dec <- "Torrejon Rubio 1"
nds.df <- read_nds(site = sit, decor = dec, dev = ".shp",
                   doss = system.file("extdata", package = "iconr"))
print(xtable::xtable(nds.df[nds.df$id == 6, ],
                     caption = caption,
                     label = "tab:nd",
                     size = 7),
      include.rownames = FALSE,
      table.placement = "!htbp")
@

@@

<<ed, results='asis', echo=TRUE >>=
caption <- paste0("First edge of the Torrejon Rubio 1 decoration, ",
                  "the proximity between the sword (node 6) ",
                  "and the fibula (node 5)")
eds.df <- read_eds(site = sit, decor = dec, dev = ".tsv",
                   doss = system.file("extdata", package = "iconr"))
print(xtable::xtable(eds.df[1,],
                     caption = caption,
                     label = "tab:ed",
                     size = 7),
      include.rownames = FALSE,
      table.placement = "!htbp")
@

@@

<<trst, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=3.5, fig.asp=1.126126, fig.cap="\\label{fig:trst}Torrejon Rubio 1 decoration">>=
par(mar=c(0, 0, 0, 0))
sit <- "Torrejon Rubio" ; dec <- "Torrejon Rubio 1"
nds.df <- read_nds(site = sit, decor = dec, dev = ".tsv",
                   doss = system.file("extdata", package = "iconr"))
eds.df <- read_eds(site = sit, decor = dec, dev = ".tsv",
                   doss = system.file("extdata", package = "iconr"))
img.graph <- plot_dec_grph(nodes = nds.df,
                           edges = eds.df,
                           site = sit,
                           decor = dec,
                           doss = system.file("extdata", 
                                              package = "iconr"),
                           nd.var = "type", lbl.size = 2)
out <- magick::image_read(img.graph)
plot(out)
@

@@

<<echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.asp=0.5630631, fig.cap="\\label{fig:ceds}Comparison between decoration 1 (Cerro Muriano 1) and decoration 4 (Zarza de Montsanchez)">>=
par(mar=c(0,0,0,0))
g.compar <- list_compar(lgrph, "type")
eds_compar <- plot_compar(g.compar, c(1, 4),
                          focus = "edges",
                          doss = system.file("extdata", package = "iconr"))
plot(magick::image_trim(magick::image_read(eds_compar)))
@

@@

<<results='asis', echo=TRUE, fig.height= 5>>=
dataDir <- system.file("extdata", package = "iconr")
nds.df <- read_nds(site = "Ibahernando",
                   decor = "Ibahernando",
                   doss = dataDir)
eds.df <- read_eds(site = "Ibahernando",
                   decor = "Ibahernando",
                   doss = dataDir)
l_dec_df <- contemp_nds(nds.df, eds.df, selected.nd = 2)
l_dec_df[[1]]$site <- l_dec_df[[1]]$decor <- NULL
l_dec_df[[2]]$site <- l_dec_df[[2]]$decor <- NULL
grph <- graph_from_data_frame(l_dec_df[[2]], vertices = l_dec_df[[1]],
                           directed = FALSE)
grph$site <- grph$decor <- "Ibahernando"
grph$name <- 5
lgrph[[5]] <- grph
@

@@

In Archaeology, the iconography has been considered as a privileged elements to the identification of cultural facies \citep{Renfrew91}. By integrating the recording, the management and the analysis process in a simple manner, within the framework of graph Theory and binary topological telationships (\emph{birel}), the \pkg{iconr} framework opens possibilities of studying graphical systems with normalized indexes over large series (Later Paleolithic cave paintings, \emph{cups-and-rings} art, Valcamonica Iron Age \emph{pittoti}, etc.) a long period of time, and on heterogeneous source data.
\\
The \pkg{iconr} framework is grounded on a parsimonious model, where Tobler's '\emph{near things}' \citep{Tobler70} of any iconographical content is recorded with a minimal of \emph{a priori} definitions. Those definitions only concern the graph elements: graphical units variables (ie, nodes variables) and types of relations they share (edges). The current edge notation with \emph{normal} edge, \emph{attribute} edge, and \emph{diachronic} edge allow to describe most cases of Prehistorical iconography, but can also be extended to accept new types like, for example, the \emph{interact} edge when two graphical units interact (ie, when an action is visible). For the nodes, we have only considered the iconographical \code{type} (\code{personnage}, \code{casque}, etc.) as the studied variable (\code{nd.var}), but the package user can create and select any other structured vocabularies: \code{technique}, \code{color}, etc. Tree structures for hierarchical datasets allow generalization processes (up, the parent level) and specification processes (down, the children level). Such a formalism can be used to weight the differences between nodes, to conduct analysis with different level of precision or to overcome idiosynchratic typologies issues. To record the relative chronology of archaeological layers (in a case of an excavation) or the relative chronology of iconographical layers (in case of a decorated support), archaeologists use quite commonly a hierachical dataset to model : the Harris matrix. The \emph{diachronic} edge, by recording the anterior-posterior relation between pairwises of graphical units, is first operator permiting to trace the Harris matrix directly of the studied decoration. The Harris matrix can be then traced with R data tree package (\pkg{data.tree}, \cite{Glur19}) and integrated into an interactive Reingold-Tilford tree diagram  (\pkg{collapsibleTree}, \cite{Khan18}).
